Random Matrix Theory Tools for the Predictive Analysis of Functional Magnetic Resonance Imaging Examinations

### 2023-05-03 paper/paper/SPIE/response_letter.tex

%%%%%% 16. Line 311: I see what they mean by "voxel time series" because of lines
%%%%%% 317-318, so it might be clearer to move those lines to directly after line 311,
%%%%%% so it is clear without having to read further.

17. The authors do a nice job of presenting the relevant results, but the
conclusions were rather general, being along the lines of "investigate the
sensitivity of choices made in the processing and analysis". That's an
important result, given previous work and shows that RMT might not "work" with
different data sets and processing workflows. The question is: what's the
takeaway: that the method is not one in which you can have a default set of
parameters that will generally give good results (like say, the recon-all
pipeline for FreeSurfer)? If RMT so sensitive to preprocessing and choice of
parameters, is this a useful method to be applied widely on fMRI data? It would
be good for this to be discussed and perhaps some recommendations given.
}

\noindent
\today, \\
Dear Dr. Gimi, \\

Thank you for your careful consideration of our manuscript
\emph{Random Matrix Theory Tools for the Predictive Analysis of Functional
Magnetic Resonance Imaging Examinations} submitted for publication in SPIE
Journal of Medical Imaging. We also thank our anonymous reviewer,
who offered highly constructive and helpful comments on the manuscript.

@@

The updated manuscript indicates changes with coloration, as suggested by
SPIE:JMI. Minor wording or other uncontroversial changes are identified in
\textcolor{orange}{orange} in the updated .pdf manuscript (or with
\verb|\textcolor{orange}{text}| in the \texttt{.tex} file). More significant
changes are identified in \textcolor{cyan}{cyan} (\verb|\textcolor{cyan}{text}|).

\begin{itemize}
\item Reviewer comments 2, 4, 5, 6, 10, 11, 13, 14, 15, 16 (see \hyperref[comments]{below}) noted easily-correctable
and uncontroversial issues relating to phrasing, clarity or ordering, and we
have made changes in line with those comments on lines 55, 115, 136, 153,
196, 224, 247, 381-387, 413 and 309, and 329-331, respectively.

@@

Our reviewer's final comment suggested (by our reading) to give our conclusion
a bit more punch. The point is well-taken, and we have reworked our conclusion
to more strongly imply a need for caution and/or carefulness in the
presentation and interpretation of RMT-fMRI-related findings.

\bigskip

\noindent
Sincerely, \\

\small
\noindent
\textbf{Corresponding Author}\\
Jacob Levman, PhD (jlevman@stfx.ca) \\
\footnotesize
\hspace*{0.25cm}St. Francis Xavier University, Department of Computer Science  \\
\hspace*{0.25cm}Martinos Center for Biomedical Imaging, Massachusetts General Hospital, Harvard Medical School (jlevman@mgh.harvard.edu) \\
\hspace*{0.25cm}Nova Scotia Health Authority, Research Affiliate (jacob.levman@nshealth.ca) \\
\normalsize

@@

% \printbibliography % biblatex
% \renewcommand{\bibsection}{}  % https://tex.stackexchange.com/questions/132646/how-to-remove-the-references-title
% \smallskip
% \bibliographystyle{mystyle}
% \bibliography{cover}  % natbib

\footnotesize
\begin{center}
\noindent\rule{6cm}{0.4pt}
\end{center}

\AtNextBibliography{\footnotesize}
\printbibliography[heading=none]

\section*{Reviewer Comments} \label{comments}
\begin{enumerate}
\item Lines 42-45: I think it's unlikely that most readers of SPIE will be
familiar with RMT and so it would be a good idea to have a bit more of an
introduction - what's a random matrix?, what does RMT try to do (e.g.,
characterize systems that can be described using matrices, statistically,
rather than deterministically.) Later they talk about what RMT has been applied
to, but it would be good to orient the reader to the general idea.

\item Line 44: "independently" should be "independent"

\item Line 56: This is a pretty broad statement, because it would seem that it's
an open question whether all complex systems would lend themselves to RMT-based
analysis.

\item Line 100: Do the authors really mean "eliminate" here? Do they instead mean
"mask", given that the information presumably doesn't disappear.

\item Line 121: "functional connectivity". Do the authors mean "functional
connectivity matrix"?

\item Line 138: This sentence is a bit garbled. Do the authors mean "often poorly
documented or even missing..."?

\item Lines 143-145: I don't think these absences raise doubts about these two
things in general, but rather that one has to carefully document the workflows
and make the data and code available to the public.

\item Line 156: Do the authors mean "allow a reasonable classification of each
scan without resorting to other information". Unclear what they mean by
"entirety".

\item Table 1: Volumes and n\_scans are not defined sufficiently. Are "Volumes"
the number of volumes per run? Is n\_scans to total number of runs? Is there
one run per participant? The participant info is in Table 2, but Table 1 should
be independently clear.

\item Line 180: "learning vs.\ rest" should be in a different font to be
consistent with the other dataset descriptions.

\item Line 208: Are the "fractional anisotropy" values mentioned here the input
variables for BET which govern how aggressive BET is and the gradient
parameter?

\item Section 3.2: While I understand that the authors are trying to make a
general statement that will apply to all features, this section is unclear
because the reader has no idea at this point what features are. Perhaps this
could be expanded.

\item Line 223: How about "computing time" and "storage space", assuming that is
what is meant.

\item Line 267: It's not clear that this section should be here, maybe at the end
of the methods section?

\item Line 292: This is a minor point, but sentences shouldn't start with an
abbreviation (E.g.) better "For example,". This also occurs in line 387.

\item Line 311: I see what they mean by "voxel time series" because of lines
317-318, so it might be clearer to move those lines to directly after line 311,
so it is clear without having to read further.

\item The authors do a nice job of presenting the relevant results, but the
conclusions were rather general, being along the lines of "investigate the
sensitivity of choices made in the processing and analysis". That's an
important result, given previous work and shows that RMT might not "work" with
different data sets and processing workflows. The question is: what's the
takeaway: that the method is not one in which you can have a default set of
parameters that will generally give good results (like say, the recon-all
pipeline for FreeSurfer)? If RMT so sensitive to preprocessing and choice of
parameters, is this a useful method to be applied widely on fMRI data? It would
be good for this to be discussed and perhaps some recommendations given.
\end{enumerate}

### 2023-05-03 paper/paper/SPIE/article.tex
update reponse letter and article to Jacob suggestions

\textcolor{orange}{When or if RMT has} real-world explanatory potential,
\textcolor{orange}{this will most likely be} when dealing with a complex system
of many (hundreds or more\cite{mehtaRandomMatrices2004}) interacting
components, \textcolor{orange}{rather than a system that is too small for
statistical regularities ro reliably surface}. If such a system has a matrix
representation, and the eigenvalues \textcolor{orange}{and spectral
observables} of this representation have distributions similar to
\textcolor{orange}{those} predicted by RMT, it suggests the system is either
highly random or chaotic. By contrast, if the observed spectra deviate
significantly from RMT-predicted spectra, this suggests otherwise. A number of
studies have used RMT to make such interpretations and comparisons between
systems\cite{santhanamStatisticsAtmosphericCorrelations2001,
jalanUncoveringRandomnessSuccess2014,
bandyopadhyayUniversalityComplexNetworks2007,
agrawalQuantifyingRandomnessProtein2014, raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003, wangSpectralPropertiesTemporal2015,
wangRandomMatrixTheory2016, matharooSpontaneousBackpainAlters2020}.

@@

RMT has also been used in ROI-based fMRI functional connectivity studies to
investigate differences between rest and task states
\cite{wangSpectralPropertiesTemporal2015}, between subjects with and without
attention-deficit hyperactive disorder (ADHD)
\cite{wangRandomMatrixTheory2016}, between pain and non-pain states
\cite{matharooSpontaneousBackpainAlters2020}, and between left-sided vs. \
right-sided motor imagery \cite{guRandomMatrixTheory2020}. While no differences
were found in the latter study, across the first three studies, the spectra of
resting or low-attention states exhibited properties close to the GOE. These
findings suggest that certain aspects of psychological processes might be
characterized, in part, by features computed from the eigenvalues of fMRI
correlation matrices, and that these features might vary in an interpretable
manner across psychological processes. If this is the case, RMT could aid in
understanding the functioning of the human brain.

@@

\textcolor{cyan}{When dealing with large inputs, any predictive analysis must
reduce that input into predictively-useful features. Any method that reduces a
larger set of data is potentially valid as a feature extraction method, and RMT,
which summarizes matrix data by summarizing the eigenvalue distributions, may
may be useful for feature extraction. However, different feature
extraction methods may have different advantages and disadvantages in terms of
the computational complexity, amount of summary or selection involved, and in
interpretability. RMT is both mathematically and computationally complex, and
thus RMT-based features should be compared to alternate features that may be
easier to understand or compute.}

@@

\textcolor{cyan}{Given the sensitivity to these decisions observed in this
study, and considering the lack of consensus regarding the preprocessing of
fMRI or other complex biological signal data, RMT should probably not be
currently considered a tool with "out-of-the-box" utility in these domains.
While RMT likely still has potential to yield insights in a variety of
contexts, our results suggest that, absent strong evidence otherwise, it is not
safe to assume that these insights will generalize broadly beyond the specific
preprocessing and analytic pipelines involved.}

>>>>

\textcolor{orange}{When or if RMT has} real-world explanatory potential,
\textcolor{orange}{this will most likely be} when dealing with a complex system
of many (hundreds or more\cite{mehtaRandomMatrices2004}) interacting
components, \textcolor{orange}{rather than a system that is too small for
statistical regularities to reliably surface}. If such a system has a matrix
representation, and the eigenvalues \textcolor{orange}{and spectral
observables} of this representation have distributions similar to
\textcolor{orange}{those} predicted by RMT, it suggests the system is either
highly random or chaotic. By contrast, if the observed spectra deviate
significantly from RMT-predicted spectra, this suggests otherwise. A number of
studies have used RMT to make such interpretations and comparisons between
systems\cite{santhanamStatisticsAtmosphericCorrelations2001,
jalanUncoveringRandomnessSuccess2014,
bandyopadhyayUniversalityComplexNetworks2007,
agrawalQuantifyingRandomnessProtein2014, raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003, wangSpectralPropertiesTemporal2015,
wangRandomMatrixTheory2016, matharooSpontaneousBackpainAlters2020}.

@@

RMT has also been used in ROI-based fMRI functional connectivity studies to
investigate differences between rest and task states
\cite{wangSpectralPropertiesTemporal2015}, between subjects with and without
attention-deficit hyperactive disorder (ADHD)
\cite{wangRandomMatrixTheory2016}, between pain and non-pain states
\cite{matharooSpontaneousBackpainAlters2020}, and between left-sided
vs.\ right-sided motor imagery \cite{guRandomMatrixTheory2020}. While no differences
were found in the latter study, across the first three studies, the spectra of
resting or low-attention states exhibited properties close to the GOE. These
findings suggest that certain aspects of psychological processes might be
characterized, in part, by features computed from the eigenvalues of fMRI
correlation matrices, and that these features might vary in an interpretable
manner across psychological processes. If this is the case, RMT could aid in
understanding the functioning of the human brain.

@@

\textcolor{cyan}{When dealing with large inputs, any predictive analysis must
reduce that input into predictively-useful features. Any method that reduces a
larger set of data is potentially valid as a feature extraction method, and RMT,
which summarizes matrix data by summarizing the eigenvalue distributions, may
be useful for feature extraction. However, different feature
extraction methods may have different advantages and disadvantages in terms of
the computational complexity, amount of summary or selection involved, and in
interpretability. RMT is both mathematically and computationally complex, and
thus RMT-based features should be compared to alternate features that may be
easier to understand or compute.}

@@

\textcolor{cyan}{Given the sensitivity to these decisions observed in this
study, and considering the lack of consensus regarding the preprocessing of
fMRI or other complex biological signal data, RMT should probably not be
currently considered a tool with "out-of-the-box" utility in these domains.
While RMT likely still has potential to yield insights in a variety of
contexts, our results suggest that, absent strong evidence otherwise, it is not
necessarily safe to assume that these insights will generalize broadly beyond
the specific preprocessing and analytic pipelines involved.}
